Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-03-19 agujournaltemplate.tex
Discussion just missing GF 90%

% HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, long-term climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
Following 2004,  there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.

We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible. With a return to higher upwelling conditions in 2014, we also see a corresponding increase in biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). 
This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

When comparing the community between the two clusters...
- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column. When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy
- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
- Discuss how similar Part 1 and 2 of Cluster 1 are.. why are they clustered together?
- Discuss why we do not observe a clear distinction in Diversity betweent he two clusters, but mention the species turnover difference and the community evenness increase.

When comparing the environmental conditions between the two clusters...

- The decrease in observed phytoplankton cells between the two clusters indicates...
% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups

@@

- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
%It is well known that the personal component has an influence on how well species are identified. Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al. 2012; Nohe et al. 2018). A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved. Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1). In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al. 2023). % from Di Cavalho et al. 2023
- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)

In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}. In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
Alterations in physical mixing affected phytoplankton dynamics,
% MEI index relevant (from 2019 REVIEW)
The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.

These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}. After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and thus the measurable diversity. We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.

@@

2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift

   The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis. Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.

@@

3.1 Turnover is different between two clusters, also corresponding to evenness between clusters

3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities

    - We group species and genus counts, to genus level, and use genus counts as variables the model predicts. This assumes that there are not very specific niches etc. within genus, but that they are homogenous --> caveat!

>>>>

This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}. These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes can not "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO". 
We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased uwpelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}). At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

When comparing the phytoplankton community between the two clusters, we see a marked reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass measured by fluorometry particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}. It should be noted that the HPLC data were analyzed by multiple labs throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} between the different laboratories. Here, we focus on the more consistently sampled fluorometric measurements. Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a}, but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.

Despite some of the difficulty in reconciling HPLC pigment data and microscopy cell counts, they agree in a reduction in diatom abundance after 2004. Cell counts of other functional groups decreased, where HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.

If we compare the two parts of the first cluster, the community did reorganize, particularly the diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.

However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}. We can see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels. 
This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
We do see a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!

--------------------------------------------------------------------------
----------
----------
----------
----------
----------

When comparing the community between the two clusters...
- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column. When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
- Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of the fluorometric chlorophyll reduction between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. 
- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy

In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}. In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
Alterations in physical mixing affected phytoplankton dynamics,

- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
- Discuss how similar Part 1 and 2 of Cluster 1 are.. Why are they clustered together?
- Discuss why we do not observe a clear distinction in Diversity between the two clusters, but mention the species turnover difference and the community evenness increase.

When comparing the environmental conditions between the two clusters...
- When looking at distributions of measured environmental variables, there is a clear trend signifying the reduction in upwelling conditions during the period from 2004 to 2014 (cluster 2).
- We have lower wind speed, lower Isotherm, higher Temp, lower Nuts (except Silicate -> river runoff) lower salinity -> less mixing, higher precipitation.

- Climate indices: AMO stark difference between clusters, cluster 1 is negative anomaly, cluster 2 is positive anomaly, corresponding to effects of AMO -> ITCZ etc.
- MEIv2 less clear, but shows a preponderance of positive anomalies in cluster 1 ... why???
% MEI index relevant (from 2019 REVIEW)
The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.

@@

- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
%It is well known that the personal component has an influence on how well species are identified. Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al. 2012; Nohe et al. 2018). A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved. Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1). In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al. 2023). % from Di Cavalho et al. 2023
- We group species and genus counts, to genus level, and use genus counts as variables the model predicts. This assumes that there are not very specific niches etc. within genus, but that they are homogenous --> caveat!
- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)

These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}. After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and thus the measurable diversity. We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.

@@

2. Mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift

   The time series includes multiple measurements of the phytoplankton community, which are bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis. Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.

@@
